---
title: "emuDB: The database format of the next iteration of the EMU speech database managment system"
author: "Raphael Winkelmann"
affiliation: "Institute Of Phonetic And Speech Processing (LMU Munich)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: vignettes.bib
vignette: >
  %\VignetteIndexEntry{autobuild tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
  
# WARNING: DOCUMENT STILL BEING WRITTEN!

# Introduction

This document describes the emuDB format that is used by the `emuR` package. This database format is meant as a simple, general purpose way of storing speech databases that may contain complex hierarchical annotations.

# DB design

The database structure is basically a set of files and folders that adhere to a certain structure and naming convention (see Figure below).

![Alt text](pics/db_structure.svg) 

The database root directory contains a single `_DBconfig.json` file which, as the name implies, contains the configuration options of the database such as its level definitions, how these levels are linked in the database hierarchy and what is displayed in the EMU-webApp. The database root folder also contains session folders (ending with `_ses`). These session folders can be used to logically group the `_bndl` folders described below.

Each session folder can contain any number of `_bndl` folders. All the files belonging to a bundle, for example a recorded sentence, are stored in the corresponding bundle folder. This includes the actual recording (`.wav`) and derived signal files in the SSFF format [@cassidy:2013a] such as formants (`.fms`) or the fundamental frequency (`.f0`), both of which can be generated using the `wrassp` package. Each bundle folder also contains the annotation file (`_annot.json`) of that bundle. This file, as the name implies, contains the actual annotations including the hierarchical linking information. JSON schema files are provided to ensure the syntactic integrity of the database (*https://github.com/IPS-LMU/EMU-webApp/tree/master/dist/schemaFiles*). The following naming restictions apply:

* the database folder has to have the same name as the `_DBconfig.json` prefix
* all session folders must be named including the suffix `_ses`. Their prefixes can be chosen by the database maintainer.
* all bundle folders must be named including suffix `_bndl`. Their prefixes can be chosen by the database maintainer.
* all files within a bundle that belong to the bundle have to have the same basename as the `_bndl` folder prefix

Files that do not follow this naming convention will simply be ignored by the database interaction functions of the `emuR` package.

Optional files that may also be included in the database root directory are the `_bundleList.json` files. These files specify which annotator is assigned to which bundles. These files are used by EMU-websocket-protocol servers that implement user management to assign the correct bundles to the annotators. The serve functions implemented in the `emuR` package DO NOT support user management which means that these files will simply be ignored.

For more detailed information about the file formats used see the "File descriptions" section of this document.
    

# Creating a emuDB

There are multiple ways of creating emuDBs. The two main strategies in creating a database are to either convert existing databases or file collection to the new format or to create new databases from scratch. Both strategies are described below.

## Converting existing databases

We currently provide a function for converting existing legacyEmuDBs to the new emuDB format. Here is an example of how to use this function:

```{r}
# load the package
library(emuR)
# get the path to a .tpl file of a legacyEmuDB that comes with the emuR package
path2extdata = system.file("extdata", package = "emuR")
tplPath = list.files(path2extdata, pattern=glob2rx('*.tpl'), full.names=TRUE, recursive = T)
# convert this legacyEmuDB to the emuDB format
convert_legacyEmuDB_to_emuDB(emuTplPath = tplPath, targetDir = tempdir(), verbose = F)
```

This will create a new emuDB in the folder provided by `tempdir()` containing all the information specified in the `.tpl` file. The name of the new emuDB is the same as the basename of the `.tpl` file from which it was generated.  

A futher function provided is the `convert_TextGridCollection_to_emuDB()` function. This function converts an existing `.TextGrid` & `.wav` file collection to the emuDB format. To be able to pair the correct files together the `.TextGrid` files as well as the `.wav` files must have the same basename. A further restriction is that the tiers contained within all the `.TextGrid` files have to be equal in name & type (equal subsets can be chosen using the `tierNames` argument of the function). As an example if all `.TextGrid` files contain the tiers `Syl: IntervalTier`, `Phonetic: IntervalTier` and `Tone: TextTier` the conversion will work, if a single `.TextGrid` of the collection has the additional tier `Word: IntervalTier` the conversion will fail but can be made to work by specifying the equal subset `equalSubset = c('Syl', 'Phonetic', 'Tone')` and passing it into the function argument `tierNames`. 

```{r}
# get the path to a folder containing .wav & .TextGrid files that come with the package
path2extdata = system.file("extdata", package = "emuR")
path2folder = file.path(path2extdata, 'legacy_emu')

# convert this TextGridCollection to the emuDB format
convert_TextGridCollection_to_emuDB(path2folder, dbName = 'newDbName', 
                                    targetDir = tempdir(), verbose = F)
```

This will create a new emuDB in the folder `tempdir()` called 'newDbName'. The emuDB will contain all the tier information of the `.TextGrid` files but without hierarchical information as `.TextGrid` files do not contain any linking information. To generate links between the generated `SEGMENTS` and `EVENTS` see the **Autobuild** section in this document.


## Creating a emuDB from scratch

To create an emuDB from scratch simply call:

```{r}
create_emuDB('fromScratchDB', tempdir(), verbose=F)
```

This will create an empty emuDB that does not have any ssffTrackDefinitions or levelsDefinitions as well as not containing any sessions or bundles. Adding these to the emuDB is described in the next section.

# Manipulating a database

The initial step in manipulating or gernerally interacting with a database is to load the according database into your current R session. People interessted in the way the emuR package caches the database are refered to the **emuDB caching** section of this document.
```{r}
# gernerate path to the emtpy fromScratchDB
dbPath = file.path(tempdir(), 'fromScratchDB')
# load database
dbName = load_emuDB(dbPath, verbose = F)
print(dbName)
```

This will load the database into it's cached form for quick access to the data. As you can see the `load_emuDB()` function returns the name of the database. This name is used to reference the loaded database in most database interaction functions of the `emuR` package. To see the currently loaded emuDBs simply call:

```{r}
list_emuDBs()
```

The emuR package contains two function to purge (==unload/remove) loaded databases from the current R session. These functions are `purge_emuDB()` and `purge_all_emuDBs()`. We will not envoke these functions here as we still wish to manipulate the loaded database. A further thing to note is that if name collitions of loaded databases occur one can still access the loaded databases using their names as well as an additional unique identifier that each emuDB has called the UUID. This UUID is also listed by the `list_emuDBs()` function.

The general function prefix naming convention of database manipulation functions for loaded databases are:

- `add_XXX` add a new instance of `XXX` to the emuDB
- `list_XXX` list the current instances of `XXX` / `get_` list the current instance of `XXX`
- `modify_XXX` modify existing instances of `XXX`
- `remove_XXX` removing existing instances of `XXX`

## Level definitions

As an example: Say we wish to add a new level definition to the already loaded empty database 'fromScratchDB'. This can be achieved by invoking the function:

```{r}
add_levelDefinition(dbName, name = 'Phonetic', type = 'SEGMENT')
```

To check if this action was successfull we can simply list the current level definitions by calling:

```{r}
list_levelDefinitions(dbName)
```

If we wish to change the name/type of the level we have just added to the database we can invoke (NOTE: If there are actual instances of items present in the emuDB of this name this will not be possible):   

```{r}
# modify
modify_levelDefinition(dbName, name = 'Phonetic', newName='phonetic')
# list
list_levelDefinitions(dbName)
```

Finally we could remove the levelDefinition with the function `remove_levelDefinition()` which we will not invoke here as we will use the 'phonetic' level below. 

## Attribute definitions


Each levelDefinition can contain multiple attributes, the most common attribute beeing a label (`"type": "STRING"`). Due to this it is possible to have multiple parallel labels in a single level. This means that a single annotation item can contain multiple labels while sharing other proporties such as the start and end segment information. This can be quite handy in modeling certain types of data. A illustrative example of this would be the 'phonetic' level created above. It is often the case that databases contain the phonetic transscript using IPA UTF-8 symbols as well as using SAMPA. This is a perfect choice for using mutliple attributeDefinitions within a single level:

```{r}
# list
# list_attributeDefinition(dbName, 'phonetic')
```

## Link definitions

Todo write

## SSFF track definitions

Todo write 

  
# Autobuilding
  
The annotation of large complex multileveled hierachical speech databases is
a very time consuming laborious procedure. The *emuR* provides functions to
automate various bits of this proccedure to help with this laborious task.

This document describes how to automatically build hierachies for a given 

**emuDB**


The available function are:

- autobuild.linkFromTimes()
- autobuild.MapLevels()
- autobuild.Syllabify()



An important prereq: linkDefinition available!


## autobuild.linkFromTimes()

This function causes annotations at two levels *L1* and *L2* to be linked whenever the time(s) of the annotations at tier
*U* are within those of ttier *T*.


Given we have a a super-level of the form:

Lets load a demo database and add a new fake level to the database so we have
something to play with


```{r}
# load the package
# library(emuR)
# get the path to the ae database that comes with the package
# path2ae = system.file("extdata/emu/DBs/ae/", package = "emuR")
# load the database
# ae = load_emuDB(path2ae, verbose = F)
# plot the phonetic level
# add linkDef.
  #tmpLinkDef = create.schema.linkDefinition(type='ONE_TO_MANY', superlevelName='Phonetic', sublevelName='Tone')
  #ae$DBconfig$linkDefinitions[[length(ae$DBconfig$linkDefinitions) + 1]] = tmpLinkDef 


```

and a sub-level of the form:

<svg xmlns="http://www.w3.org/2000/svg" version="1.1">
  <rect width="100%" height="75" stroke-width="1" fill="#E7E7E7" stroke="rgb(0, 0, 0)" />
  <line x1="0" y1="0" x2="20%" y2="20%" fill="rgb(0, 0, 0)" stroke-width="5"  />
</svg>

- ONE\_TO\_MANY
- MANY\_TO\_MANY
- ONE\_TO\_ONE

# File descriptions

## _DBconfig.json

The `DBconfig` file, as mentioned above, contains the configuration options of the database. People familiar with the legacy EMU system will recognize this to be the replacement file for the legacy template (`.tpl`) file. As is the convention in quite a few programming languages variables / strings written entierly in capital letters indicate a constant variable that usually has a special meaning. This is also the case with strings like this found in the `DBconfig` (`"STRING"`,`"ITEM"`,`"SEGMENT"`, `"EVENT"`, ... ).

The `_DBconfig.json` file contains the following fields:

* `"name"` specifying the name of the database
* `"UUID"` a unique ID given to each database
* `"mediafileExtension"` the main mediafileExtension (*currently only uncompressed mono 16-bit `.wav` files are supported in every component of the EMU system. This is also the recommended audio format for the EMU system.*)
* `"ssffTrackDefinitions"` an array of definitions defining the SSFF tracks of the database. Each ssffTrackDefinition consists out of:
    * `"name"` the name of the ssffTrackDefinition
    * `"columnName"` the name of the column of the associated SSFF file. For more information on the columns the various function of the `wrassp` produce see the track field of `wrasspOutputInfos` object that is part of the `wrassp` package. Further, although the SSFF file format is a binary file format it has a plain text header which means that if you open a SSFF file with the text editor of your choice you will be able to see the columns contained within it. A further option is to get column information about a specific SSFF file, is to use the `wrassp` function `res = read.AsspDataObj('/path/2/SSFF/file')` to read the file from the file system. `names(res)` will then give you the names of the columns present in this file. NOTE: In the context of the SSFF file format the term *column* and in the context of the EMU system the term *track*/*ssffTrack* is used. They both refer to the same data.
    * `"fileExtention"` the file extension of the associated SSFF file (also see `?wrasspOutputInfos` for the default extensions produced by the `wrassp` functions)
* `"levelDefinitions"`  array of definitions defining the levels of the database. A level is a more general term for what is often refered to as a tier. It is more general in the sense that people quite often associate tiers to contain time information. Levels can however either contain time information if they are of are either of the type `"EVENT"` or of the type `"SEGMENT"` but can also be timeless if they are of the type `"ITEM"`. Each "`levelDefinitions"` consists out of:
    * `"name"` the name of the levelDefinition
    * `"type"` specifying the type of the level (either `"ITEM"` | `"EVENT"` | `"SEGMENT"`)
    * `"attributeDefinitions"` an array of definitions defining the attributes of the level. Each attributeDefinition consists out of:
        * `"name"` the name of the `"attributeDefinition"`
        * `"type"` specifying the type of the attribute (currently only `"STRING"` permitted)
        * `"labelGroups"` an array containing label group definitions. These can be used as a shorthand notation for querying certain groups of labels.
            * `"name"` name of label group. This will be the value used in a query to refer to this group 
            * `"values"` array of strings representing the labels
        * `"legalLabels"` array of strings specifying which labels are valid/legal for this attribute definition. The EMU-webApp adheres to this set of values and will not let the annotator enter any other values than the ones specified in this field. This can be used to insure consistent label sets within levels.
    * `"anagestConfig"` if specified (optional) this will convert the level into a special type of level for labeling articulatory data. This will also envoke the EMU-webApp to treat this level differently. It is optional field may only be set for levels of type `"EVENT"`. For more information on handeling articulatory data in the EMU system see XXX.
        * `"verticalPosSsffTrackName"` name of ssffTrack containing the vertical position data
        * `"velocitySsffTrackName"` name of ssffTrack containing the velocity data
        * `"autoLinkLevelName"` name of level that will be used to link the created events to
        * `"multiplicationFactor"` factor to multiply with (either `-1` | `1`)
        * `"threshold"` a value between 0 and 1 defining the threshold
        * `"gestureOnOffsetLabels"` array containing two strings that specify the on- and offset labels
        * `"maxVelocityOnOffsetLabels"` array containing two strings that specify the on- and offset labels
        * `"constrictionPlateauBeginEndLabels"` array containing two strings that specify the begin- and end labels
        * `"maxConstrictionLabel"` string specifying label
* `"linkDefinitions"`  an array of definitions defining the levels of the database. The combination of all link definition specifies the hierarchy of the database. Each ssffTrackDefinition consists of:
    * `"type"` specifying the type of link (either `"ONE_TO_MANY"` | `"MANY_TO_MANY"` | `"ONE_TO_ONE"`).
    * `"superlevelName"` specifies the name of the superlevel
    * `"sublevelName"` specifies the name of the sublevel
* `"labelGroups"` an array containing label group definitions. These can be used as a shorthand notation for querying certain groups of labels. Compared to the `"labelGroups"` that can be defined within a attributeDefinition the labelGroups defined here are globally defined for the entire database.
    * `"name"` name of label group. This will be the value used in a query to refer to this group 
    * `"values"` array of strings containing con
* `"EMUwebAppConfig"` specifies the configuration of the EMU-webApp to display the database. As the EMU-webApp has multiple configuration options please refer to it's manual (TODO ADD LINK!) for more information about it's configuration options.


# emuDB caching

TODO: Write about object relational mapping into sqlite DB 
