---
title: "An introduction to the emuR package"
subtitle: "The main package of the EMU Speech Database Management System"
affiliation: "Institute Of Phonetics And Speech Processing (LMU Munich)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: vignettes.bib
vignette: >
  %\VignetteIndexEntry{emuR introduction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Introduction

# Converting existing databases

The `emuR` package provides a function for converting existing legacyEmuDBs to the new emuDB format. Here is an example of how to use this function:

```{r}
# load the package
library(emuR)
# create demo data in folder provided by the tempdir() function
create_emuRdemoData()
# get the path to a .tpl file of a legacyEmuDB that is part of the demo data
tplPath = file.path(tempdir(), "emuR_demoData", "legacy_ae", "ae.tpl")
# convert this legacyEmuDB to the emuDB format
convert_legacyEmuDB_to_emuDB(emuTplPath = tplPath, targetDir = tempdir(), verbose = F)
```

This will create a new emuDB in a temporary folder that is provided by the `R` function `tempdir()` containing all the information specified in the `.tpl` file. The name of the new emuDB is the same as the basename of the `.tpl` file from which it was generated. In other words, if the template file of your legacyEmuDB has the path `A` and the directory to which the converted database is to be written has the path `B`, then  `convert_legacyEmuDB_to_emuDB("A", targetdir = "B")` will perform the conversion.

A further function provided is the `convert_TextGridCollection_to_emuDB()` function. This function converts an existing `.TextGrid` & `.wav` file collection to the emuDB format. In order to  pair the correct files together the `.TextGrid` files as well as the `.wav` files must have the same name (i.e. file name without extension). A further restriction is that the tiers contained within all the `.TextGrid` files have to be equal in name & type (equal subsets can be chosen using the `tierNames` argument of the function). For example, if all `.TextGrid` files contain the tiers `Syl: IntervalTier`, `Phonetic: IntervalTier` and `Tone: TextTier` the conversion will work. However, if a single `.TextGrid` of the collection has the additional tier `Word: IntervalTier` the conversion will fail, although it can be made to work by specifying the equal subset `equalSubset = c('Syl', 'Phonetic', 'Tone')` and passing it into the function argument `convert_TextGridCollection_to_emuDB(..., tierNames = equalSubset, ...)`. 

```{r}
# get the path to a folder containing .wav & .TextGrid files that is part of the demo data
path2folder = file.path(tempdir(), "emuR_demoData", "TextGrid_collection")

# convert this TextGridCollection to the emuDB format
convert_TextGridCollection_to_emuDB(path2folder, dbName = 'myTGcolDB', 
                                    targetDir = tempdir(), verbose = F)
```

This will create a new emuDB in the folder `tempdir()` called 'myTGcolDB'. The emuDB will contain all the tier information from the `.TextGrid` files but will not contain hierarchical information as `.TextGrid` files do not contain any linking information. To generate links between the generated `SEGMENTS` and `EVENTS` see the [Autobuilding](#autobuilding) section of this document.


# extracting / manipulating and visualizing data



```{r}
# cleanup (probably better to do it directly after you created it)
unlink(file.path(tempdir(),'emuR_demoData'), recursive = T)
unlink(file.path(tempdir(),'ae'), recursive = T)
unlink(file.path(tempdir(),'myTGcolDB'), recursive = T)
```

